### 2025-01-24 skripsi/apendix.tex
add example tensors for pre and post processing, add SoTA section

>>>>

\chapter{EXAMPLE GENERATION AND PROCESSING OF DATA BEFORE AND AFTER THE LSSVR}\label{ch:example_prepostprocessing_basis}
First we will start with how a function can be generated randomly. The random solution and parameter pair generation process starts by randomly sampling from a normal distribution coefficients of Fourier basis functions which result in tensors as shown in \lccref{fig:example_generated_coefficients}. The dimensions of this tensor of coefficients is the number of unique functions \(N\) by the number of basis functions representing each dimension \(m_1, m_2, \ldots, m_d\).
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
tensor([[ 0.0000+0.0000j,  5.9584-7.5454j, -6.4740-6.0901j,  ...,
          4.5666+7.3829j, -6.4740+6.0901j,  5.9584+7.5454j],
        [ 0.0000+0.0000j,  5.4649-6.3776j, -4.8105+1.3955j,  ...,
          4.8338-1.8225j, -4.8105-1.3955j,  5.4649+6.3776j],
        [ 0.0000+0.0000j,  1.4697-5.8133j,  7.7991+2.8807j,  ...,
          2.0915-1.7939j,  7.7991-2.8807j,  1.4697+5.8133j],
        ...,
        [ 0.0000+0.0000j, -0.7932+4.6427j,  6.5799+1.1859j,  ...,
         -0.1577-9.3086j,  6.5799-1.1859j, -0.7932-4.6427j],
        [ 0.0000+0.0000j,  5.2905-6.0955j, -5.6544+3.8172j,  ...,
          5.8754-1.5761j, -5.6544-3.8172j,  5.2905+6.0955j],
        [ 0.0000+0.0000j,  3.2358+6.8016j, -8.0814+3.5778j,  ...,
         -5.1760+1.9684j, -8.0814-3.5778j,  3.2358-6.8016j]])
\end{lstlisting}
  \caption{Example of coefficients randomly generated as the solution of the 1D derivative equation.}\label{fig:example_generated_coefficients}
\end{figure}

Noise is then added to the generated coefficients by transforming the coefficients to their function values using the function shown in \lccref{fig:fourier_transform_impl}. For our previous tensor of randomly generated coefficients the inverse transform results in the tensor shown in \lccref{fig:example_clean_inv_transform}. This tensor has the shape of \(N\) unique functions by the number of points in each dimension \(n_1,n_2,\ldots,n_d\).
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
  tensor([[-1.4411-4.4703e-08j,  0.3935-2.2054e-08j, -1.4868-8.9565e-09j,
          ..., -0.7727-4.2093e-08j, -0.3239+2.0381e-08j,
          1.1053-8.6315e-08j],
        [-0.9293-5.7220e-08j, -0.0500-4.7684e-08j, -0.2628+2.8610e-08j,
          ...,  1.4031-5.8692e-08j,  0.1245-6.1988e-08j,
          2.1139+2.9045e-08j],
        [-0.7640-2.8610e-08j,  1.0191+6.1989e-08j,  0.8502+2.7388e-08j,
          ..., -1.1099-5.4252e-08j,  0.3513+2.3798e-08j,
          1.0349-1.5378e-07j],
        ...,
        [ 0.3640+1.7881e-09j, -0.6330+5.5432e-08j, -1.6459-1.1356e-07j,
          ...,  1.9432+7.8670e-08j,  0.1409+1.5608e-08j,
          1.9429-1.1898e-07j],
        [ 1.4188-5.7220e-08j, -0.0971-1.9073e-08j,  1.2511-6.4455e-09j,
          ..., -0.2919+8.5157e-08j, -0.2904+1.2012e-08j,
          0.4568-1.1858e-07j],
        [-1.1443-9.5367e-09j, -1.6120+9.0599e-08j,  0.7427-3.5344e-08j,
          ..., -0.6559+4.0621e-08j,  0.6636+2.1595e-08j,
          0.7892-6.0481e-08j]])
\end{lstlisting}
  \caption{Example of the inverse Fourier transform of the randomly generated coefficients.}\label{fig:example_clean_inv_transform}
\end{figure}

The values we have obtained are perturbed using random values sampled from a normal distribution and then scaled to the requested noise ratio based on the standard deviation of the computed function values. Once perturbed, the values are transformed back to the Fourier domain.

For data that starts as function values, like measurements or the ERA5 dataset, this is where the computational process would start. The function values transformed into the coefficient representation is shown in \lccref{fig:example_data_fourier_transformed}. This tensor has the same shape as \lccref{fig:example_generated_coefficients}. For data that started as function values, the number of modes would be by default the same as the discretization in each dimension. This can be altered according to the problem.
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
tensor([[ 7.1488e-01-1.0658e-14j,  6.4083e+00-7.7222e+00j,
         -6.8440e+00-6.9812e+00j,  ...,
          4.9350e+00+6.8183e+00j, -6.8440e+00+6.9812e+00j,
          6.4083e+00+7.7222e+00j],
        [ 1.4488e-01-9.5923e-14j,  5.7198e+00-6.4293e+00j,
         -5.0511e+00+1.5238e+00j,  ...,
          4.6795e+00-1.7510e+00j, -5.0511e+00-1.5238e+00j,
          5.7198e+00+6.4293e+00j],
        [-9.2841e-01+6.3949e-14j,  2.2419e+00-5.2065e+00j,
          7.8745e+00+2.6073e+00j,  ...,
          1.3050e+00-8.7435e-01j,  7.8745e+00-2.6073e+00j,
          2.2419e+00+5.2065e+00j],
        ...,
        [-1.1377e-03+1.4211e-14j,  2.9627e-01+4.8204e+00j,
          7.0653e+00+7.0965e-01j,  ...,
         -1.1042e-01-9.8381e+00j,  7.0653e+00-7.0965e-01j,
          2.9627e-01-4.8204e+00j],
        [-8.5599e-01-6.0396e-14j,  4.9924e+00-5.7394e+00j,
         -6.8187e+00+3.7167e+00j,  ...,
          5.8396e+00-1.1584e+00j, -6.8187e+00-3.7167e+00j,
          4.9924e+00+5.7394e+00j],
        [ 7.6012e-01-2.1316e-14j,  3.6402e+00+5.9960e+00j,
         -7.6304e+00+3.8345e+00j,  ...,
         -4.6177e+00+2.4993e+00j, -7.6304e+00-3.8345e+00j,
          3.6402e+00-5.9960e+00j]])
\end{lstlisting}
  \caption{Example of the coefficients from the Fourier transform of the data which was either generated and perturbed or from external sources as function values like ERA5 2-meter temperature.}\label{fig:example_data_fourier_transformed}
\end{figure}

For the generated data, the entire process of adding noise and getting the coefficients of the perturbed function values is done using the perturb function of the base Basis class as shown in \lccref{fig:basis_perturb_impl}.
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
...
    def perturb(
        self,
        std_ratio: float = 0.1,
        rand_func=torch.randn,
        generator: torch.Generator | None = None,
    ) -> Self:
        values = self.inv_transform(self.coeff)
        perturbed_values = (
            values
            + rand_func(
                values.shape,
                generator=generator,
                dtype=self.coeff.dtype
                if self._complex_funcs
                else self.coeff.real.dtype,
            )
            * std_ratio
            * values.std()
        )
        copy = self.copy()
        copy.coeff = self.transform(perturbed_values)
        return copy
...
\end{lstlisting}
  \caption{Example of the inverse Fourier transform of the randomly generated coefficients.}\label{fig:basis_perturb_impl}
\end{figure}

Finally, the tensor of noisy generated functions or function values from other sources are preprocessed to be able to be used with the LSSVR\@. First, since LSSVR only works with one dimensional feature vectors, we flatten the tensors so that they become matrices. The coefficients are complex types, we represent them as two elements of the float type in the tensor as shown in \lccref{fig:example_coeff_flat_real}. Notice that the values are just the components of the complex number in \lccref{fig:example_data_fourier_transformed} as their own elements side by side. This would have the shape of \(N\) unique functions by the product of the number of modes in each dimension \(m_1\times m_2 \times\cdots\times m_d\) and multiplied by two because of the representation of complex number components as real elements. The final shape is \(N\) by \(2\times m_1\times m_2 \times\cdots\times m_d\).
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
tensor([[ 7.1488e-01, -1.0658e-14,  6.4083e+00,  ...,  6.9812e+00,
          6.4083e+00,  7.7222e+00],
        [ 1.4488e-01, -9.5923e-14,  5.7198e+00,  ..., -1.5238e+00,
          5.7198e+00,  6.4293e+00],
        [-9.2841e-01,  6.3949e-14,  2.2419e+00,  ..., -2.6073e+00,
          2.2419e+00,  5.2065e+00],
        ...,
        [-1.1377e-03,  1.4211e-14,  2.9627e-01,  ..., -7.0965e-01,
          2.9627e-01, -4.8204e+00],
        [-8.5599e-01, -6.0396e-14,  4.9924e+00,  ..., -3.7167e+00,
          4.9924e+00,  5.7394e+00],
        [ 7.6012e-01, -2.1316e-14,  3.6402e+00,  ..., -3.8345e+00,
          3.6402e+00, -5.9960e+00]])
\end{lstlisting}
  \caption{Example of the flattened coefficients represented as real numbers side by side.}\label{fig:example_coeff_flat_real}
\end{figure}

For the real representation of the parameter functions which will act as features, the data is scaled to have a distribution with a mean of zero and standard deviation of zero. This is the standard scaling of the data to achieve better and more consistent performance with the LSSVR\@. To achieve this, the current mean and standard deviation of the data is computed for each feature column (coefficient component). By subtracting the current data by the computed mean and then dividing by the standard deviation, the data should now have a mean of roughly zero and standard deviation of one. For our example data the scaled version is shown in \lccref{fig:example_scaled_data}.
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
tensor([[ 1.0025, -0.0103,  1.2830,  ...,  1.4096,  1.2830,  1.5487],
        [ 0.1999, -0.0955,  1.1464,  ..., -0.3085,  1.1464,  1.2894],
        [-1.3114,  0.0643,  0.4562,  ..., -0.5274,  0.4562,  1.0442],
        ...,
        [-0.0057,  0.0146,  0.0701,  ..., -0.1440,  0.0701, -0.9666],
        [-1.2095, -0.0600,  1.0020,  ..., -0.7515,  1.0020,  1.1511],
        [ 1.0662, -0.0209,  0.7337,  ..., -0.7753,  0.7337, -1.2024]])
  \end{lstlisting}
  \caption{Example of the scaled parameter function coefficients real representation.}\label{fig:example_scaled_data}
\end{figure}

The scaled parameter function coefficients which will act as the features along with the labels of the solution function coefficients are then used to get the trained parameters of the LSSVR\@. Since the predictions will be in the same format as shown in \lccref{fig:example_coeff_flat_real}, to obtain the predicted function values, we simply do the reverse process until we get to the function values. First the coefficients are changed to their complex number representation and then the dimensions are unflattened which results in a similar format to \lccref{fig:example_data_fourier_transformed}. So for labels that were originally two-dimensional, this means the predictions need to be reshaped from the flattened state into the same two-dimensional form. So the flattened dimension with size \(2\times m_1\times m_2 \times\cdots\times m_d\) would become \(d\) dimensions with shape \(m_1, m_2, \ldots, m_d\). Once the coefficients are formatted correctly, we use the inverse Fourier transform to obtain the function values with the implementation of \lccref{eq:inverse_discrete_fourier_transform}. The resulting function values should have the same format as the example shown in \lccref{fig:example_clean_inv_transform}. Since for the most part we are only interested in real-valued functions, we can discard the imaginary component. This results in the final predicted solution function values.

### 2025-01-24 skripsi/chapter2.tex
update outline to be like the UPI format with chapter abstracts
add example tensors for pre and post processing, add SoTA section

% !TEX root = ./skripsi.tex
\chapter{LITERATURE REVIEW}\label{sec:literature_review}
% \noindent Bab ini mengulas secara rinci konsep-konsep dasar yang berkaitan dengn pekerjaan penelitian TA dan deskripsi studi pustaka yang dilakukan. Judul bab tidak harus seperti yang dituliskan, melainkan dapat lebih fleksibel yang mencerminkan isi paparan pada bab ini. Demikian halnya dengan judul sub bab.
\begin{figure}[H]
  \begin{adjustwidth}{-0.1\linewidth}{-0.1\linewidth}
    \centering
    \tikzfig{figures/literature_map}
  \end{adjustwidth}
  \caption{Literature Map}\label{fig:literature_map}
\end{figure}
The topics this literature review will cover in this chapter is shown in \lccref{fig:literature_map}. There are four main topics which are governing equations, Fourier transform and series, machine learning regression, and relevant works to this study. These topics will be discussed in the following sections.
\section{Governing Equations}
\noindent Mathematical expressions that describe the behavior of a system are called governing equations. Typically, governing equations are in the form of Differential Equations (DE). For some systems with several variables, the governing equations are Partial Differential Equations (PDE). These equations are the foundation of many engineering and scientific models \autocite{olverIntroductionPartialDifferential2014,andersonGoverningEquationsFluid1992}. In the physical sciences, the use of PDEs as governing equations is widespread. Many fields, including those outside the physical sciences, use these equations to model systems and solve problems whose solutions we take for granted today, ranging from fluid mechanics to ecology.

@@

There are currently different model architectures for the learning operators. Most of these are implemented as Neural Networks (NN). The Neural Operator proposes a method to generalize NNs to learning mappings between function spaces \autocite{li2021fourier}. The Fourier Neural Operator (FNO) variant adds Fourier transforms to the composition of layers within the NN\@. Another NN layer is added to learn the mapping within the Fourier domain. The outputs of this layer is then Inverse transformed and passed on. This entire process is called the FNO block. Adding multiple of these blocks allow for learning more complex relationships. Another approach called Spectral Neural Operators, instead maps the coefficients of inputs to the coefficients of output functions \autocite{fanaskovSpectralNeuralOperators2023}. For coefficients obtained from Fourier transforms, the model is approximating the standard spectral method by taking advantage of the properties of the Fourier coefficients we discuss in \lccref{sec:fourier_discussion}. Essentially, derivatives are multiplication in the Fourier domain. A similar work extended the idea by leveraging the known PDE to be part of the loss function \autocite{du2024neural}. The model is penalized for the residual computed using the PDE\@. This allows the NN to be trained in a self-supervised setting.

Another paradigm is the so called DeepONet architecture proposed by \textcite{luLearningNonlinearOperators2021}. This model can be seen as foregoing the basis function the Fourier Transform enforces and instead learns the basis functions from the data \autocite{meurisMachinelearningbasedSpectralMethods2023}. Because of this, the model has two component networks. The first, named the trunk network acts as the learned basis functions. This network is given evaluation coordinates and computes the values of the \enquote{basis functions} at said coordinates. The second network, called the branch network, learns the mapping between the input function representation and the coefficients for the output function. The dot product of the outputs of both trunk and branch networks are evaluated to obtain the value of the predicted function at the specified coordinate. The final proposed approach we will discuss here doesn't use Deep Learning. The approach proposed by \textcite{nelsenOperatorLearningUsing2024} uses Random Feature Models (RFM) to learn operators. This model enjoys the advantage of using one of the simplest forms of machine learning.

>>>>

% !TEX root = ./skripsi.tex
\chapter{LITERATURE REVIEW}\label{ch:literature_review}
% \noindent Bab ini mengulas secara rinci konsep-konsep dasar yang berkaitan dengn pekerjaan penelitian TA dan deskripsi studi pustaka yang dilakukan. Judul bab tidak harus seperti yang dituliskan, melainkan dapat lebih fleksibel yang mencerminkan isi paparan pada bab ini. Demikian halnya dengan judul sub bab.
\begin{figure}[H]
  \begin{adjustwidth}{-0.1\linewidth}{-0.1\linewidth}
    \centering
    \tikzfig{figures/literature_map}
  \end{adjustwidth}
  \caption{Literature Map}\label{fig:literature_map}
\end{figure}
The topics this literature review will cover in this chapter is shown in \lccref{fig:literature_map}. There are four main topics which are governing equations, Fourier transform and series, machine learning regression, and relevant works to this study. These topics will be discussed in the following sections.
\section{Governing Equations}
\noindent Mathematical expressions that describe the behavior of a system are called governing equations. Typically, governing equations are in the form of Differential Equations (DE). For some systems with several variables, the governing equations are Partial Differential Equations (PDE). These equations are the foundation of many engineering and scientific models \autocite{olverIntroductionPartialDifferential2014,andersonGoverningEquationsFluid1992}. In the physical sciences, the use of PDEs as governing equations is widespread. Many fields, including those outside the physical sciences, use these equations to model systems and solve problems whose solutions we take for granted today, ranging from fluid mechanics to ecology.

@@

There are currently different model architectures for the learning operators. Most of these are implemented as Neural Networks (NN). The Neural Operator proposes a method to generalize NNs to learning mappings between function spaces \autocite{li2021fourier}. The Fourier Neural Operator (FNO) variant adds Fourier transforms to the composition of layers within the NN\@. Another NN layer is added to learn the mapping within the Fourier domain. The outputs of this layer is then Inverse transformed and passed on. This entire process is called the FNO block. Adding multiple of these blocks allow for learning more complex relationships. Another study extends this to spherical harmonics which reduce artifacts when modeling spherical domains like the weather \autocite{bonevSphericalFourierNeural2023}. % TODO: add more and maybe move some of the text in the introduction here.

\section{State of The Art}
\noindent Within the State of The Art (SoTA), our proposed method falls within a group of operator learning approach that utilize the advantageous properties of basis functions. This group of methods exploit the same properties of basis functions and their coefficients that the more conventional spectral method also utilize. One of the earliest works in this group of methods is the Deep Operator Network (DeepONet) \textcite{luLearningNonlinearOperators2021}. This paradigm is different from later methods since it uses a neural network as a learned basis function. This model can be seen as foregoing the basis function the Fourier transform enforces and instead learns the basis functions from the data \autocite{meurisMachinelearningbasedSpectralMethods2023}. Because of this, the model has two component networks. The first, named the trunk network acts as the learned basis functions. This network is given evaluation coordinates and computes the values of the \enquote{basis functions} at said coordinates. The second network, called the branch network, learns the mapping between the input function representation and the coefficients for the output function. The dot product of the outputs of both trunk and branch networks are evaluated to obtain the value of the predicted function at the specified coordinate.

Another approach called Spectral Neural Operators, instead maps the coefficients of more common basis functions like the Fourier or Chebyshev basis functions \autocite{fanaskovSpectralNeuralOperators2023}. For coefficients obtained from Fourier transforms, the model is approximating the standard spectral method by taking advantage of the properties of the Fourier coefficients we discuss in \lccref{sec:fourier_discussion}. Essentially, derivatives are multiplication in the Fourier domain. A similar work extended the idea by leveraging the known PDE to be part of the loss function \autocite{du2024neural}. The model is penalized for the residual computed using the PDE\@. This allows the NN to be trained in a self-supervised setting. The final proposed approach we will discuss here doesn't use Deep Learning. The approach proposed by \textcite{nelsenOperatorLearningUsing2024} uses Random Feature Models (RFM) to learn operators. This model enjoys the advantage of using one of the simplest forms of machine learning.

### 2025-01-24 skripsi/chapter4.tex
add example tensors for pre and post processing, add SoTA section

Another important utility used on inputs for the proposed SpectralSVR is a scaling function. This utility has been implemented in a variety of ways and using different tools. However, as the standard scaling is not provided by PyTorch, we have also implemented our own version. The scaling function accepts PyTorch tensors or tuples of tensors of real or complex elements. First, the scaling function is implemented as a class that is instantiated. The scaling function is then fitted onto the reference data to obtain the standard deviation and mean of each feature. This information is stored as properties of the class. Any complex tensors are converted into their real representations using the functions in \lccref{fig:complex_matrix_conversion_implementation}. Once the scaling function instance is fitted, any set of tensors with the same number of features (columns) and element type (complex or real) can be roughly transformed into a standard deviation of one and mean of zero for every feature. This is done by subtracting the fitted mean of each feature from the elements in the column and then dividing with the feature's standard deviation. The implementation also provides a method to retrieve a scaling function fitted on a subset of the original tuple of tensors. Serialization for saving to and loading from disk is also implemented. This is done simply by using the PyTorch save and load functions on the scaling function instance itself. For some use cases, an inverse scaling operation to the original standard deviation and mean is also essential. The implementation is simply the reverse of the transformation process where the elements are multiplied by the standard deviation and the mean is the added.

@@

\noindent This module implements two classes. The first is the base Basis class. The second is a subclass implementing the Fourier basis specifically. These classes provide the functionality needed to have an easier time when working with basis functions and their coefficients. First, the core function of these classes is to provide a way to transition between the spectral domain and the physical domain. This is implemented in the base class by storing the coefficients of sample functions as a class property. The coefficients are assumed to be at least of one function.

>>>>

Another important utility used on inputs for the proposed SpectralSVR is a scaling function. This utility has been implemented in a variety of ways and using different tools. However, as the standard scaling is not provided by PyTorch, we have also implemented our own version. The scaling function accepts PyTorch tensors or tuples of tensors of real or complex elements. First, the scaling function is implemented as a class that is instantiated. The scaling function is then fitted onto the reference data to obtain the standard deviation and mean of each feature. This information is stored as properties of the class. Any complex tensors are converted into their real representations using the functions in \lccref{fig:complex_matrix_conversion_implementation}. Once the scaling function instance is fitted, any set of tensors with the same number of features (columns) and element type (complex or real) can be roughly transformed into a standard deviation of one and mean of zero for every feature. This is done by subtracting the fitted mean of each feature from the elements in the column and then dividing with the feature's standard deviation. The implementation also provides a method to retrieve a scaling function fitted on a subset of the original tuple of tensors. Serialization for saving to and loading from disk is also implemented. This is done simply by using the PyTorch save and load functions on the scaling function instance itself. For some use cases, an inverse scaling operation to the original standard deviation and mean is also essential. The implementation is simply the reverse of the transformation process where the elements are multiplied by the standard deviation and the mean is the added. This functionality is demonstrated for processing the data for use in SpectralSVR in \lccref{ch:example_prepostprocessing_basis}.

@@

\noindent This module implements two classes. The first is the base Basis class. The second is a subclass implementing the Fourier basis specifically. These classes provide the functionality needed to have an easier time when working with basis functions and their coefficients. An example of the use of these functions for pre-processing and post-processing the function data into and out of the Fourier domain is presented in \lccref{ch:example_prepostprocessing_basis}. First, the core function of these classes is to provide a way to transition between the spectral domain and the physical domain. This is implemented in the base class by storing the coefficients of sample functions as a class property. The coefficients are assumed to be at least of one function.
